---
title: "Data Mining course 2022/2023"
subtitle: "Lab1 - a short introduction to R (with examples)"
author: Andrzej Janusz
email: ap.janusz@uw.edu.pl
output:
  html_notebook:
    df_print: paged
    fig_height: 8
    fig_width: 10
    rows.print: 10
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, results=FALSE}
# list of libraries used in this script
library(data.table)
library(microbenchmark)

```

### The plan:

0.  Organizational matters.
1.  Using *help*.
2.  Operations on numbers, vectors, and matrices.
3.  Basic data types and classes.
4.  Vector/matrix indexing methods.
5.  Basic statistics.
6.  Basic programming constructs.
7.  Function definition, passing parameters.

### Work organization and assessment rules

We will use Moodle:
<https://moodle.mimuw.edu.pl/course/view.php?id=1767> The registration
key: 0!41UWrl

To pass the laboratories, you will have to get a sufficient number of
points. You will receive points for:

-   *active* participation in classes (up to $10$ points),
-   points for solving two projects (up to $2 \times 25 = 50$ points).

Additional points can be awarded for solving additional tasks (not
mandatory) or for active participation in a data mining competition (up
to 20 points). Exemplary (currently ongoing) challenge:
<https://knowledgepit.ai/icme-2023-grand-challenge/>

The total number of collected points from laboratories can not exceed
60. Points collected during the laboratories will influence the final
grade (together with points from the exam).

### Useful links:

R project web page:\
<http://www.r-project.org/>

Resources for learning R:

-   *Introduction to R*\
    <http://cran.r-project.org/doc/manuals/R-intro.html>
-   *R Language* reference manual\
    <http://cran.r-project.org/doc/manuals/R-lang.html>
-   A nice tutorial to the *data.table* package\
    <https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html>

```{r help, eval= FALSE}
# you start R by typing
R

# and you end an R session with
q()

# you can read manual to a specific function by typing a question mark '?' before its name
?q

# you can also browse manuals of all installed R functions and libraries by typing
help.start()
```

### Arithmetics:

```{r vectors, eval = FALSE}
# but you need to remember that in R even a single number is a vector...
is(1)

# so, vector arithmetics is the same as the arithmetics on numbers
c(1,2,3,4,5) # - this is a concatenation of five vectors...
(1:10) - 1
(1:10) * 2
2^(1:10)
(1:10)^2
log(1:10)
(1:10)*(10:1)

# some methods of defining simple vectors
1:10
rep(1, 10)
rep(1:2, 5)
rep(1:2, each = 5)

seq(1,10,by=1)
seq(5,10,length.out = 10)
rep(1:5, seq(2,10, by=2))

# defining 'empty' vectors:
integer(0)
character(10)
numeric(10)
```

```{r matrices, eval = FALSE}
# methods of creating matrices
1:10 %*% 10:1
1:10 %*% t(1:10)

matrix(1:100, nrow = 10, ncol = 10)
matrix(1:100, nrow = 10, byrow=TRUE)
matrix(rep(1:5, seq(2,10, by=2)), nrow = 10)

diag(1:10)
diag(1:10 %*% t(1:10))

# in R, a matrix is also a vector
myMatrix = matrix(c(1,2,3,3,2,1,1,2,1), nrow = 3, byrow=TRUE)
myMatrix
is(myMatrix)

# simple arithmetics:
myMatrix * myMatrix
myMatrix + 1
myMatrix + 1:3
myMatrix * 2
myMatrix * 1:3

solve(myMatrix)
solve(myMatrix, c(2,3,1))
myMatrix %*% solve(myMatrix, c(2,3,1))
```

### Operators and functions

```{r operators, eval = FALSE}
# comparison operators: <, <=, >, >=, ==, !=
1 < 2
1:10 > 5
1:10 == c(1:3, 6:4, 7:10)

# various types of assignment operators =, <-, <<-, ->, ->>
x = 4
y <- 5
6 -> z
c(x, y, z)

'='(z, 1)
z

# small differences between '=' and '<-' operators
x; mean(1:10); x

mean(x = 1:10); x

mean(x <- 1:10); x

# mean(abc = 1:10) would be an error (check it), but
mean(abc <- 1:10) # what happens here?
abc

# some more advanced example
new_counter <- function() {
  i <- 0
  function() {
    # here you can add some important piece of code for which you want to count executions
    i <<- i + 1 # double arrow looks at the env 'one up', each function has its own enviroment
    i
  }
}

i <- 10

counter_one <- new_counter()
counter_two <- new_counter()

counter_one() 
counter_one()

counter_two()

i

# second example (as a warning)
test <- function() {
  mySum <- 0
  for (i in c(1, 2, 3)) { # for doesn't change the enviroment, so the mySum is from outside
    mySum <<- mySum + i 
  }
  mySum
}

test()

mySum # what happened here?

# logical operators !, &, |, &&, ||, xor(), any(), all(), which()
x = 2 > 1
x & 3 < 4
!x | 5^5 > 5*5
xor(x, 5^5 > 5*5)
any(1:10 > 5)
all(1:10 > 5)
which(1:10 != c(1:3, 6:4, 7:10))
```

### Atomic data types and basic data structures

```{r data_types, eval = FALSE}
# in R, atomic data types are: logical, integer, numeric, complex, character, raw
# basic data structures: vectors and lists

# classes of R objects
class(1:10)
is(1:10)
is(as.numeric(1:10))
class("abc")

# dynamic typing in R
class(1:3)
x = c(1:3, 4.5)
class(x)
x[7] = "d"
x
class(x)

y = 1234
as.character(y)

# naming elements of vectors
names(x)
names(x) = c("a", "b", "c", "D", "", "", "X")
names(x)

x[2]
x["X"]
x["bleble"]

# factors in R
myFactor = factor(rep(c("A", "B"), 5), levels = c("A", "B", "C"))
is(myFactor)
levels(myFactor)
as.integer(myFactor)
as.character(myFactor)
```

```{r lists_and_frames, eval = FALSE}
# creating a list
myList = list(vector = x, label = y)
myList

myList$vector
myList$vec
myList$v
myList[[2]] # chooses the second element
myList[2]   # chooses a list that stores the second element
class(myList[[2]])
class(myList[2])

# data frames are the basic data storage in R (not the most efficient)
myDataFrame = data.frame(0)
myDataFrame = as.data.frame(myMatrix)
is(myDataFrame)
as.list(myDataFrame)

# you can get a list of all exemplary data frames available in loaded libraries by typing 'data()'
myDataFrame = iris
head(myDataFrame)
dim(myDataFrame)
nrow(myDataFrame)
ncol(myDataFrame)
colnames(myDataFrame)

# adding columns and rows
myDataFrame = cbind(myDataFrame, myDataFrame[,1] * myDataFrame[,2] - 1)
# you need to remember to check the column types
myDataFrame = rbind(myDataFrame, myDataFrame[1:3,])
```

```{r indexing, eval = FALSE}
# in R, you can index data using vectors of integers (indexes), logical values, names or expressions
myVector = seq(100, 0, -2)
myVector[26:35]
myVector[1:length(myVector) %in% 26:35]

# modifying vectors
mySecondVector = character(length(myVector))
mySecondVector[myVector %% 4 == 0] = "class1"
mySecondVector[mySecondVector != "class1"] = "class2"
head(mySecondVector)

# indexes do not need to be unique
myDataFrame[c(1,1,2,3,1,1,2),]
```

### Some basic programming constructs

```{r statistics, eval = FALSE}
# in R, you can find ready-to-use implementations of many types functions commonly used in statistics
# for example: abs, sum, min, max, pmin, pmax, quantile, median, prod, mean, rank, order, sort, ....
myVector  = seq(10, 1, length.out = 7)
sum(myVector)
min(myVector)
rank(myVector)
mean(myVector)
sort(myVector)
max(myVector, sort(myVector))
pmax(myVector, sort(myVector))
```

```{r applys, eval = FALSE}
# a method to iterate over various data structures - 'apply' functions
apply(myDataFrame[1:4], 1, sum)
apply(myDataFrame[1:4], 2, mean)

sapply(myDataFrame[1:4], mean)
lapply(myDataFrame[1:4], median)

# there are many other versions of *apply functions, each dedicated to a different data type
```

```{r conds, eval = FALSE}
# conditional instructions:
if(_condition_) {}
else {}
```

```{r conds_cd, eval = FALSE}
if(mean(myDataFrame$Sepal.L) > min(myDataFrame$Sepal.W)) {
  print("L > W") 
} else {
  print("L <= W")
}
```

```{r loops, eval = FALSE}
# loops - use them only if it is really necessary

# 'for' loop:
for(name in _expression1_) {_expression2_}

# 'while' loop:
while(_condition_) {_expression_}

# 'infinite' loop: 
repeat _expression_
```

```{r functions}
# definition of functions:
myFunction = function(x, y, f, w = 0, ...) { # this ... parameter is really interesting...
    f(x^2 - y^2 + w, ...)
}

myFunction(3,2, function(x, a = 0) x + a)

myFunction(3,2, function(x, a = 0) x + a, a = 100)

# recursion:
multiply = function(x,y,w = 0)  {
  if(y==0) w 
  else multiply(x, y-1 ,w+x)
}

multiply(4,5)

```

### Data reading and writing

There are plenty tools for I/O opperations in R.

```{r data_read_write, eval = FALSE}
# Predefined data sets in loaded R libraries:
data()                # the list of currently available data sets

# a simple data editor:
fix(myData) # do not use it on large data sets!
# a simple data viewer:
View(myData)

# R has tools for importing and exporting data from majority of popular data formats 
# http://cran.r-project.org/doc/manuals/R-data.pdf

# basic functions to read the data:
?scan
?read.table
?readLines
?data.table::fread

# tools for importing data from other popular data exploration systems (e.g., EpiInfo, Minitab, S-PLUS, SAS, SPSS, Stata, Systat)
# in the "foreign" library

# APIs to various data bases: e.g., libraries RMySQL, ROracle, RPostgreSQL, RSQLite

# writing matrices/data frames/data tables to files
write.table(myData, file = "iris1.data", sep="\t", col.names=TRUE, row.names=FALSE)
write.csv(myData, file = "iris2.data")
write.csv2(myData, file = "iris3.data")

# writing data to binary (compressed) files for fast read/write:
save(myData, iris, file = "someData.RData") # it can be used to serialize a pre-trained model
save.image(file = "allData.RData")

# clearing memory
rm(myData)
gc()       # you can also run the garbage collector manually

# loading binary files
load(file = "someData.RData")
```

### Using the data.table package

The *data.table* package powerful tools for in-memory processing of
large data sets.

```{r data_read, echo=FALSE, warning=FALSE, message=FALSE}
# some very useful libraries:
library(parallel)
library(data.table)
library(dplyr)
library(tidyr)

# typically, we load all libraries in the setup section of the notebook

# A part of the data.table tutorial from the package's vignettes:
# data reading
input <- if (file.exists("flights14.csv")) {
   "flights14.csv"
} else {
  "https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv"
}
flights <- fread(input)

# we save the csv file so we don't need to download it again
fwrite(flights, file = "flights14.csv")
```

```{r data_processing_examples}
head(flights)

tracemem(flights)
# fast data querying:
microbenchmark::microbenchmark(
  ans <- flights[origin == "JFK" & month == 6L]
)
head(ans)

microbenchmark::microbenchmark(
  ans2 <- flights[order(origin, -dest)]
)
head(ans2)

# subsetting and aggregating
microbenchmark::microbenchmark(
  ans3 <- flights[origin == "JFK" & month == 6L,
                  .(m_arr = mean(arr_delay), m_dep = mean(dep_delay))]
)
ans3

# you can efficiently run 'group by' queries:
microbenchmark::microbenchmark(
  ans4 <- flights[carrier == "AA",
                  list(mean(arr_delay), mean(dep_delay)),
                  keyby = .(origin, dest, month)]
)
ans4

# applying aggregations to multiple columns
microbenchmark::microbenchmark(
  ans5 <- flights[carrier == "AA",                       ## Only on trips with carrier "AA"
                  lapply(.SD, mean),                     ## compute the mean
                  by = .(origin, dest, month),           ## for every 'origin,dest,month'
                  .SDcols = c("arr_delay", "dep_delay")] ## for just those specified in .SDcols
)
ans5
```

```{r dplyr_vs_data.table}
# a quick comparison with dplyr (from tidyverse)
flights <- as_tibble(flights) # we convert our data.table to a tibble to make things eraser for dplyr
is(flights)

microbenchmark::microbenchmark(
  ans <- filter(flights,
                (origin == "JFK") & (month == 6L))
)
head(ans)
```

```{r dplyr_vs_data.table_cd, message=FALSE}
# we may turn the memory-reporting off to negate its impact
untracemem(flights)
microbenchmark::microbenchmark(
  ans <- filter(flights,
                (origin == "JFK") & (month == 6L))
)

# now, a 'group by' query
microbenchmark::microbenchmark(
  ans4 <- flights %>%
          filter(carrier == "AA") %>%
          group_by(origin, dest, month) %>%
          summarise(mean_arr_delay = mean(arr_delay), mean_dep_delay = mean(dep_delay))
)
ans4
```

### Exercises

```{r task1}
# Example 1:
vec1 <- rnorm(10000000)
vec2 <- rnorm(10000000)
vec3 <- numeric(10000000)

rep_num <- 5

# compare execution times of multiplying in a loop and using vectors
microbenchmark::microbenchmark(
  {
    for(i in 1:length(vec1)) vec3[i] <- vec1[i] * vec2[i]
  },
  times = rep_num
)
print(sum(vec3))

microbenchmark::microbenchmark(
  {
    vec3 <- vec1 * vec2
  },
  times = rep_num
)
print(sum(vec3))
```

```{r task2}
# Example 2: for a given numeric matrix or a data.frame compute a difference between 
# the largest and the smallest element in each column, without using a loop.
DT <- iris

max_min_diff <- function(vec) max(vec) - min(vec)

sapply(DT[, 1:4], max_min_diff)
```

```{r task3}
# Example 3:
dim(iris)
DT <- data.table::rbindlist(replicate(1000000, iris, simplify = FALSE))

is(DT)
dim(DT)
object.size(DT)

colnames(DT)

system.time({
  print(DT[, list(mean.Sepal.Length = mean(Sepal.Length),
                  mean.Sepal.Width = mean(Sepal.Width)), 
           by = Species])
})

data.table::setkey(DT, 'Species')
system.time({
  print(DT[, list(mean.Sepal.Length = mean(Sepal.Length),
                  mean.Sepal.Width = mean(Sepal.Width)), 
           by = Species])
})

rm(DT)
```

**Exercise 1:** You have a numeric vector of length 1000000. Compute
sums of each three consecutive elements of this vector:\
\* using a *for* loop \* without any loop (in a vectorized manner)\
Compare the execution times.

**Exercise 2:** You have a list of size 1000. Each element of that list
is a numeric vector of length 100. Convert this list into a 1000 x 100
matrix:\
\* using a *for* loop and the *rbind* function\
\* in a vectorized manner.\
Compare the execution times.

**Exercise 3:** Use the *data.table* package to compute dominant values
for all symbolic (character) columns in the *Flights* data set.

```{r the_end}
# clearing the memory and manually running the garbage collector
rm(list = ls()); gc()
```
